Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-09-10 writeup/pnas-word-order-si.tex

We also collected, for each language, all auxiliaires, that is, all dependents of the \emph{aux} dependency, occurring at least 10 times, and compared their dominant orders.
In X languages, all extracted auxiliaries had the same dominant order.
In X languages, there were auxiliaires with divergent order. In Y cases, these were modal auxiliaries, or uninflected elements. In Z cases, these were tense-aspect auxiliaries.
Taken together, this shows that the dominant order of the \emph{aux} relation strongly coincides with that of inflected tense-aspect or negation auxiliaries.

@@

@@

>>>>

We also collected, for each language, all auxiliaires, that is, all dependents of the \emph{aux} dependency, occurring at least 10 times, and compared their dominant orders, which we operationalized as their more common order in the treebank (auxiliary-head or head-auxiliary).
In X languages, all extracted auxiliaries had the same dominant order.
In X languages, there were auxiliaires with different dominant orders.
		In Y cases, these were modal or passive auxiliaries, or uninflected elements.
		In Z cases, these were tense-aspect auxiliaries:
		In Persian and Afrikaans, not all tense-aspect auxiliaries 
		(1) In Persian, not all tense-aspect auxiliaries have the same order: E.g. budan as a perfect auxiliary follows the main verb (Mace p. 117), whereas xaastan xaah- as a future auxiliary precedes it (Mace p. 122).
		(1) In Afrikaans, the past auxiliary \emph{het} shows dominant aux-head order, whereas the overall most common order of auxiliaries is head-aux (e.g. future auxiliary \emph{sal}).
		(2) In Old Church Slavonic, most aux dependencies have the reflexive \emph{sebe} which tends to precede the head, whereas inflected verbal auxiliaries (primarily, forms of \emph{byti}) have opposite order, leading to a seeming violation of the correlation  \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}.

Taken together, this shows that the dominant order of the \emph{aux} relation strongly coincides with that of inflected tense-aspect or negation auxiliaries.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized.pdf}
	\caption[Predictability and Parseability]{Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}
\end{figure}

@@

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized.tex}
\end{tabular}
}
	\caption{Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}
\end{table}
